---
title: "Specs for Rhys's App"
author: "Homer White"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
knitr::opts_chunk$set(
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

## Aim of the App

The purpose of this app is to investigate the coverage properties of confidence intervals for $\mu$, the mean of a population, at various sample sizes and with consideration of a case when the population is not normally-distributed.

## Setup

```{r}
########################################
#  Utility Functions for Generating Pareto Values
########################################

rpareto <- function(n,alpha,theta) {#random values for Pareto(alpha,theta) distribution
  theta*((1-runif(n))^(-1/alpha)-1)
}

dpareto <- function(x,alpha,theta) {  #pdf for Pareto(alpha,theta) distribution
  alpha*theta^alpha/(x+theta)^(alpha+1)
}


#############################################
# Generate the populations
############################################
mu_norm <- 70
sigma_norm <- 5
shape_gamma <- 2
scale_gamma <- 50

# for pareto:
alpha_pareto <- 5
theta_pareto <- 100
tail_prob <- 0.02  #want to find a Value at risk of 1 - this
val_risk <- theta_pareto*(tail_prob^(-.5)-1)

# for pop with group of outliers
prop_outliers <- 0.10
mean_outliers <- 200
sd_outliers <- 5
mean_regulars <- 50
sd_regulars <- 5

r_outlier<- function(n) {
  prop_normals <- 1- prop_outliers
  which_hump <- rbinom(n, size = 1, prob = prop_normals)
  outlier_samp <- 
    ifelse(
      which_hump,
      rnorm(n,mean=mean_regulars,sd=sd_regulars),
      rnorm(n,mean=mean_outliers,sd=sd_outliers)
    )
  outlier_samp
}

d_outlier <- function(x) {
  reg <- dnorm(x,mean=mean_regulars,sd=sd_regulars)
  out <- dnorm(x,mean=mean_outliers,sd=sd_outliers)
  (1-prop_outliers) * reg + prop_outliers * out
}

#######################################
# Get the population means
######################################

normal_mean <- mu_norm
skew_mean <- shape_gamma*scale_gamma
super_skew_mean <- theta_pareto/(alpha_pareto - 1)
outlier_mean <- (1-prop_outliers)*mean_regulars+prop_outliers*mean_outliers
```

Function to plot a desired population:


```{r}
draw_pop <- function(pop = c("normal", "skew", "super_skew", "outlier")) {
  if (pop == "normal") {
    df <- data.frame(
      x = c(mu_norm - 3 * sigma_norm, mu_norm + 3 * sigma_norm)
    )
    p <-
      ggplot(df, aes(x = x)) +
      stat_function(
        fun = dnorm,
        n = 101,
        args = list(
          mean = mu_norm,
          sd = sigma_norm
        )
      ) + 
      labs(
        x = "population value",
        y = NULL
      ) +
      geom_vline(aes(xintercept = normal_mean))
  }
  if (pop == "skew") {
    df <- data.frame(
      x = c(0, shape_gamma*scale_gamma+7.5*sqrt(shape_gamma)*scale_gamma)
    )
    p <-
      ggplot(df, aes(x = x)) +
      stat_function(
        fun = dgamma,
        n = 101,
        args = list(
          shape = shape_gamma,
          scale = scale_gamma
        )
      ) + 
      labs(
        x = "population value",
        y = NULL
      ) +
      geom_vline(aes(xintercept = skew_mean))
  }
  if (pop == "super_skew") {
    df <- data.frame(
      x = c(0, val_risk)
    )
    p <-
      ggplot(df, aes(x = x)) +
      stat_function(
        fun = dpareto,
        n = 101,
        args = list(
          alpha = alpha_pareto,
          theta = theta_pareto
        )
      ) + 
      labs(
        x = "population value",
        y = NULL
      ) +
      geom_vline(aes(xintercept = super_skew_mean))
  }
  if (pop == "outlier") {
    df <- data.frame(
      x = c(0, mean_outliers +5 * sd_outliers)
    )
    p <-
      ggplot(df, aes(x = x)) +
      stat_function(
        fun = d_outlier,
        n = 201
      ) + 
      labs(
        x = "population value",
        y = NULL
      ) +
      geom_vline(aes(xintercept = outlier_mean))
  }
  p
}
```

Try it out:

```{r}
pop <- "outlier"
draw_pop(pop)
```

Hmm, should the mean-line be a different color, and or different width?

Here is a function to get m samples, each of size n:


```{r}
get_samples <- function(m, n, pop = c("normal", "skew", "super_skew", "outlier")) {
  samps <- vector(mode = "list", length = m)
  for (i in 1:m) {
    if (pop == "normal") {
      samps[[i]] <- rnorm(n, mean = normal_mean, sd = sigma_norm)
    }
    if (pop == "skew") {
      samps[[i]] <- rgamma(n, shape = shape_gamma, scale = scale_gamma)
    }
    if (pop == "super_skew") {
      samps[[i]] <- rpareto(n, alpha = alpha_pareto, theta = theta_pareto)
    }
    if (pop == "outlier") {
      samps[[i]] <- r_outlier(n)
    }
  }
  samps
}
```

Try it out:

```{r}
get_samples(m = 3, n = 20, pop = "outlier")
```


Functions to compute sample means and confidence intervals, and keep track of how many cover the mean:

```{r}
get_intervals <- function(samps, pop = c("normal", "skew", "super_skew", "outlier"), level) {
  if (pop == "normal") mu <- normal_mean
  if (pop == "skew") mu <- skew_mean
  if (pop == "super_skew") mu <- super_skew_mean
  if (pop == "outlier") mu <- outlier_mean
  m <- length(samps)
  good <- logical(m)
  xbar <- numeric(m)
  lower <- numeric(m)
  upper <- numeric(m)
  number = 1:m
  for (i in 1:m) {
    xs <- samps[[i]]
    n <- length(xs)
    xbar[i] <- mean(xs)
    crit <- qt((1 + level) / 2, df = n-1)
    margin <- crit * sd(xs) / sqrt(n)
    lower[i] <- xbar[i] - margin
    upper[i] <- xbar[i] + margin
    good[i] <- mu >= lower[i] & mu <= upper[i]
  }
  data.frame(
    number = number,
    xbar = xbar,
    lower = lower,
    upper = upper,
    good = good
  )
}
```

Try it out:

```{r}
pop <- "skew"
some_samples <- get_samples(m = 3, n = 20, pop = pop)
level <- 0.90
get_intervals(
  samps = some_samples,
  pop = pop,
  level = level
)
```


Function to draw your intervals:

```{r}
interval_plot <- function(data, pop = c("normal", "skew", "super_skew", "outlier")) {
  if (pop == "normal") mu <- normal_mean
  if (pop == "skew") mu <- skew_mean
  if (pop == "super_skew") mu <- super_skew_mean
  if (pop == "outlier") mu <- outlier_mean
  plot <- ggplot(data) +
    geom_vline(aes(xintercept = mu)) +
    geom_segment(
      aes(
        x = lower,
        xend = upper,
        y = number,
        yend = number,
        color = good
      )
    ) +
    geom_point(aes(x = xbar, y = number), size = 0.5) +
    labs( x = NULL, y = NULL) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    )
  plot
}
```

Try it out:

```{r}
pop <- "skew"
some_samples <- get_samples(m = 50, n = 20, pop = pop)
level <- 0.90
ints <- get_intervals(
  samps = some_samples,
  pop = pop,
  level = level
)
interval_plot(data = ints, pop = pop)
```



## App Specs

* Sidebar Panel
    * select input, id = "pop", values are `c("normal", "skew", "super_skew", "outlier")`
    * numeric input, id = "n" (to determine sample size)
    * slider input, id = "level" (desired confidence level)
    * radio button input, id = "m" (number of intervals to draw in simulation; choices are 50 and 100)
    * action button with id = "make_intervals"
    * action button with id = "start_over"
* Main Panel (tabset with two tabs)
    * first tab has two rows:
        * first row is a plot of the population
        * second row states the mean $\mu$ of the population
    * second tab has two rows:
        * first row: plot showing either the requested 50 or 100 confidence intervals for the proportion
        * second row: statement of total intervals so far, number of interval that contain the proportion, and percentage of intervals that contains the proportion.

When the user pushes the Make Intervals button, the pop, sample size, and level inputs are hidden.  Fifty or 100 new samples are drawn.  Confidence intervals are computed and in the second tab the new plot appears and the statement underneath is updated to reflect the new total number of intervals made so far.

When the user pushes the Start Over button, the four hidden widgets reappear, the record of the intervals is destroyed, the graph on the first row is hidden and the second-row statement indicates that there are no intervals yet.

**Note**: you should think about how to modify the `interval_plot()` function so that the intervals and point-estimates are still easy to see with 100 intervals.

Once you have the app working, you might want to think about how to improve it.  For example, could the user be allowed to say how many samples she wants at once, with a minimum of 50, and the interval plot shows just the final 50 intervals in the sample.  If so, what should be the maximum allowed number of samples at once?  Is there are way to speed up the makiing of samples and the computation of intervals so that this maximum can be set quite high?

